---
title: "Introduction to taxdiv"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to taxdiv}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The **taxdiv** package provides tools for calculating taxonomic diversity
indices that incorporate the hierarchical structure of biological
classification. While traditional diversity indices like Shannon and Simpson
only consider species abundances, taxonomic diversity measures account for
the evolutionary and ecological relationships among species.

The package implements three main approaches:

1. **Classical diversity indices**: Shannon and Simpson
2. **Clarke & Warwick taxonomic distinctness**: Delta, Delta*, AvTD, VarTD
3. **Ozkan (2018) Deng entropy-based diversity**: pTO and pTO+

```{r setup}
library(taxdiv)
```

## Example Data: Mediterranean Forest Community

We will use a hypothetical Mediterranean forest community to demonstrate the
package functionality. This community has 10 tree and shrub species with
known abundances and a 4-level taxonomic hierarchy.

```{r data}
# Species abundances (individuals per plot)
community <- c(
  Quercus_coccifera    = 25,
  Quercus_infectoria   = 18,
  Pinus_brutia         = 30,
  Pinus_nigra          = 12,
  Juniperus_excelsa    = 8,
  Juniperus_oxycedrus  = 6,
  Arbutus_andrachne    = 15,
  Styrax_officinalis   = 4,
  Cercis_siliquastrum  = 3,
  Olea_europaea        = 10
)

# Taxonomic hierarchy
tax_tree <- build_tax_tree(
  species = names(community),
  Genus   = c("Quercus", "Quercus", "Pinus", "Pinus",
              "Juniperus", "Juniperus", "Arbutus", "Styrax",
              "Cercis", "Olea"),
  Family  = c("Fagaceae", "Fagaceae", "Pinaceae", "Pinaceae",
              "Cupressaceae", "Cupressaceae", "Ericaceae", "Styracaceae",
              "Fabaceae", "Oleaceae"),
  Order   = c("Fagales", "Fagales", "Pinales", "Pinales",
              "Pinales", "Pinales", "Ericales", "Ericales",
              "Fabales", "Lamiales")
)

tax_tree
```

## Classical Diversity Indices

Shannon and Simpson indices measure diversity based solely on species
abundance distribution:

```{r classical}
# Shannon diversity (natural log)
H <- shannon(community)
cat("Shannon H':", round(H, 4), "\n")

# Simpson indices
D <- simpson(community, type = "dominance")
GS <- simpson(community, type = "gini_simpson")
inv_D <- simpson(community, type = "inverse")

cat("Simpson dominance (D):", round(D, 4), "\n")
cat("Gini-Simpson (1-D):", round(GS, 4), "\n")
cat("Inverse Simpson (1/D):", round(inv_D, 4), "\n")
```

These indices tell us about the evenness and richness of the community, but
they do not consider that some species are taxonomically more distant from
each other than others.

## Taxonomic Distance

Before computing taxonomic diversity, we can examine the pairwise
taxonomic distances between species:

```{r distance}
dist_mat <- tax_distance_matrix(tax_tree)
round(dist_mat, 2)
```

Species in the same genus (e.g., *Quercus coccifera* and *Q. infectoria*)
have distance 0 at all shared taxonomic levels, while species in different
orders have the maximum distance.

## Clarke & Warwick Taxonomic Distinctness

The Clarke & Warwick framework provides abundance-weighted and
presence/absence-based measures:

```{r cw}
# Delta: average taxonomic diversity (abundance-weighted)
d <- delta(community, tax_tree)
cat("Delta (taxonomic diversity):", round(d, 4), "\n")

# Delta*: taxonomic distinctness (abundance-weighted, excludes same-species)
ds <- delta_star(community, tax_tree)
cat("Delta* (taxonomic distinctness):", round(ds, 4), "\n")

# AvTD (Delta+): average taxonomic distinctness (presence/absence)
spp <- names(community)
avg_td <- avtd(spp, tax_tree)
cat("AvTD (Delta+):", round(avg_td, 4), "\n")

# VarTD (Lambda+): variation in taxonomic distinctness
var_td <- vartd(spp, tax_tree)
cat("VarTD (Lambda+):", round(var_td, 4), "\n")
```

## Deng Entropy and Ozkan pTO

The Deng entropy framework (Deng, 2016) generalizes Shannon entropy through
Dempster-Shafer evidence theory. At each taxonomic level, the mass function
accounts for the number of species within each group (the "focal element
size"), giving more weight to groups that contain more species.

Ozkan (2018) uses Deng entropy to construct four measures:

- **uTO**: Unweighted taxonomic diversity (uses slicing procedure)
- **TO**: Weighted taxonomic diversity (taxonomic levels weighted by rank)
- **uTO+**: Unweighted taxonomic distance (presence/absence, nk=0 only)
- **TO+**: Weighted taxonomic distance

```{r pto}
result <- ozkan_pto(community, tax_tree)

cat("uTO  (unweighted diversity):", round(result$uTO, 4), "\n")
cat("TO   (weighted diversity):", round(result$TO, 4), "\n")
cat("uTO+ (unweighted distance):", round(result$uTO_plus, 4), "\n")
cat("TO+  (weighted distance):", round(result$TO_plus, 4), "\n")
```

The Deng entropy values at each taxonomic level reveal the contribution
of each hierarchical level to overall diversity:

```{r ed_levels}
cat("Deng entropy at each level:\n")
for (i in seq_along(result$Ed_levels)) {
  cat("  ", names(result$Ed_levels)[i], ":",
      round(result$Ed_levels[i], 4), "\n")
}
```

At the species level, Ed equals ln(S) = ln(10) since all present species
receive equal weight. At higher levels, the Deng correction term
(2^|Fi| - 1) increases entropy for groups containing more species.

### Convenience wrapper

The `pto_components()` function returns all four values as a named vector:

```{r components}
pto_components(community, tax_tree)
```

## Comparing Communities

Taxonomic diversity measures are most useful when comparing communities.
Here we compare our Mediterranean community with a species-poor community:

```{r compare}
# Degraded community (fewer species, less taxonomic breadth)
degraded <- c(
  Quercus_coccifera   = 40,
  Pinus_brutia        = 35,
  Juniperus_oxycedrus = 10
)

tax_degraded <- tax_tree[tax_tree$Species %in% names(degraded), ]

cat("=== Original community (10 species) ===\n")
cat("Shannon:", round(shannon(community), 4), "\n")
r1 <- ozkan_pto(community, tax_tree)
cat("uTO+:", round(r1$uTO_plus, 4), "\n")
cat("TO+:", round(r1$TO_plus, 4), "\n\n")

cat("=== Degraded community (3 species) ===\n")
cat("Shannon:", round(shannon(degraded), 4), "\n")
r2 <- ozkan_pto(degraded, tax_degraded)
cat("uTO+:", round(r2$uTO_plus, 4), "\n")
cat("TO+:", round(r2$TO_plus, 4), "\n")
```

The degraded community shows lower diversity across all measures, but the
taxonomic indices (uTO+, TO+) capture not only the loss of species richness
but also the narrowing of taxonomic breadth.

## Stochastic Resampling (Run 2)

The deterministic pTO calculation (Run 1) uses all species as they are.
But how sensitive is the result to community composition? Run 2 explores
this by randomly including or excluding each species with 50% probability
across many iterations, then taking the maximum pTO value observed.

```{r run2}
set.seed(42)
run2 <- ozkan_pto_resample(community, tax_tree, n_iter = 101, seed = 42)

cat("=== Run 1 (deterministic) ===\n")
cat("uTO+:", round(run2$uTO_plus_det, 4), "\n")
cat("TO+: ", round(run2$TO_plus_det, 4), "\n")
cat("uTO: ", round(run2$uTO_det, 4), "\n")
cat("TO:  ", round(run2$TO_det, 4), "\n\n")

cat("=== Run 2 (max across", run2$n_iter, "iterations) ===\n")
cat("uTO+:", round(run2$uTO_plus_max, 4), "\n")
cat("TO+: ", round(run2$TO_plus_max, 4), "\n")
cat("uTO: ", round(run2$uTO_max, 4), "\n")
cat("TO:  ", round(run2$TO_max, 4), "\n")
```

The maximum values from Run 2 are always >= the deterministic values from
Run 1, because the deterministic calculation is included as the first
iteration.

We can examine how the pTO values vary across iterations:

```{r run2_summary}
iter_df <- run2$iteration_results
cat("uTO+ range:", round(min(iter_df$uTO_plus), 4), "to",
    round(max(iter_df$uTO_plus), 4), "\n")
cat("TO+  range:", round(min(iter_df$TO_plus), 4), "to",
    round(max(iter_df$TO_plus), 4), "\n")
```

## Sensitivity Analysis (Run 3)

Run 3 refines the exploration by assigning species-specific inclusion
probabilities based on the Run 2 results. Species that contributed to
higher diversity in Run 2 get different inclusion probabilities than
those that did not.

```{r run3}
run3 <- ozkan_pto_sensitivity(community, tax_tree, run2, seed = 123)

cat("=== Run 3 (sensitivity analysis) ===\n")
cat("Run 3 max uTO+:", round(run3$run3_uTO_plus_max, 4), "\n")
cat("Run 3 max TO+: ", round(run3$run3_TO_plus_max, 4), "\n\n")

cat("=== Overall max (across Run 1 + 2 + 3) ===\n")
cat("uTO+:", round(run3$uTO_plus_max, 4), "\n")
cat("TO+: ", round(run3$TO_plus_max, 4), "\n")
cat("uTO: ", round(run3$uTO_max, 4), "\n")
cat("TO:  ", round(run3$TO_max, 4), "\n")
```

The overall maximum across all three runs represents the "potential"
taxonomic diversity of the community — the highest diversity that can
be observed under different species compositions derived from the
original community.

### Species Inclusion Probabilities

Run 3 assigns each species a probability of being included in each
iteration:

```{r species_probs}
probs <- run3$species_probs
prob_df <- data.frame(
  Species = names(probs),
  Probability = round(probs, 4)
)
print(prob_df, row.names = FALSE)
```

## Full Pipeline Summary

The complete Ozkan (2018) analysis pipeline runs three stages:

```{r pipeline_summary}
cat("Pipeline: Run 1 -> Run 2 -> Run 3\n\n")

cat("         uTO+      TO+       uTO       TO\n")
cat("Run 1:", sprintf("%9.4f %9.4f %9.4f %9.4f",
    run2$uTO_plus_det, run2$TO_plus_det,
    run2$uTO_det, run2$TO_det), "\n")
cat("Run 2:", sprintf("%9.4f %9.4f %9.4f %9.4f",
    run2$uTO_plus_max, run2$TO_plus_max,
    run2$uTO_max, run2$TO_max), "\n")
cat("Run 3:", sprintf("%9.4f %9.4f %9.4f %9.4f",
    run3$uTO_plus_max, run3$TO_plus_max,
    run3$uTO_max, run3$TO_max), "\n")
```

Each subsequent run finds values >= the previous run, reflecting
the increasing exploration of the diversity landscape.

## References

- Deng, Y. (2016). Deng entropy. *Chaos, Solitons & Fractals*, 91, 549-553.
- Ozkan, K. (2018). A new proposed measure for estimating taxonomic
  diversity. *Turkish Journal of Forestry*, 19(4), 336-346.
- Clarke, K.R. & Warwick, R.M. (1998). A taxonomic distinctness index and
  its statistical properties. *Journal of Applied Ecology*, 35, 523-531.
- Warwick, R.M. & Clarke, K.R. (1995). New 'biodiversity' measures reveal
  a decrease in taxonomic distinctness with increasing stress. *Marine
  Ecology Progress Series*, 129, 301-305.
