---
title: "Ozkan pTO Method: Deng Entropy-Based Taxonomic Diversity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ozkan pTO Method: Deng Entropy-Based Taxonomic Diversity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

## What Is the Ozkan pTO Method?

Ozkan (2018) introduced a novel approach to measuring taxonomic diversity
using **Deng entropy** --- a generalization of Shannon entropy rooted in
Dempster-Shafer evidence theory (Dempster, 1967; Shafer, 1976). Ozkan &
Mert (2022) extended this with **reinforced estimators** that use stochastic
resampling (Run 2) and sensitivity analysis (Run 3) to refine the indices.

The key idea: at each level of the taxonomic hierarchy (genus, family,
order, etc.), **Deng entropy** measures how evenly species are distributed
across groups. The product of these level-wise entropies gives a single
number that captures the **entire hierarchical diversity** of a community.

This approach produces **8 complementary indices** through a three-stage
pipeline, each answering a slightly different question about the community.

```{r setup}
library(taxdiv)

community <- c(
  Quercus_coccifera    = 25,
  Quercus_infectoria   = 18,
  Pinus_brutia         = 30,
  Pinus_nigra          = 12,
  Juniperus_excelsa    = 8,
  Juniperus_oxycedrus  = 6,
  Arbutus_andrachne    = 15,
  Styrax_officinalis   = 4,
  Cercis_siliquastrum  = 3,
  Olea_europaea        = 10
)

tax_tree <- build_tax_tree(
  species = names(community),
  Genus   = c("Quercus", "Quercus", "Pinus", "Pinus",
              "Juniperus", "Juniperus", "Arbutus", "Styrax",
              "Cercis", "Olea"),
  Family  = c("Fagaceae", "Fagaceae", "Pinaceae", "Pinaceae",
              "Cupressaceae", "Cupressaceae", "Ericaceae", "Styracaceae",
              "Fabaceae", "Oleaceae"),
  Order   = c("Fagales", "Fagales", "Pinales", "Pinales",
              "Pinales", "Pinales", "Ericales", "Ericales",
              "Fabales", "Lamiales")
)
```

## From Shannon to Deng: Why a New Entropy?

Shannon entropy treats each species as an independent event with probability
$p_i$. But in a taxonomic hierarchy, species are grouped --- two oak species
share more information than an oak and a pine. Shannon cannot capture this
grouping.

**Deng entropy** solves this through the concept of **focal elements** from
evidence theory. At each taxonomic level, a group (e.g., "Family Fagaceae")
acts as a focal element with a mass proportional to the species it contains.
The entropy accounts for both the mass distribution and the **size of each
focal element** (how many species it contains):

$$E_d = -\sum_{i=1}^{n} m(F_i) \log_2 \frac{m(F_i)}{2^{|F_i|} - 1}$$

where $m(F_i)$ is the mass of focal element $F_i$ and $|F_i|$ is the number
of species it contains.

The term $2^{|F_i|} - 1$ accounts for all possible non-empty subsets of
species within the group. A genus with 3 species has $2^3 - 1 = 7$ possible
subcombinations, giving it more "evidential weight" than a single-species
genus.

## Deng Entropy at Each Taxonomic Level

```{r deng_levels}
result <- ozkan_pto(community, tax_tree)

cat("Deng entropy by taxonomic level:\n\n")
for (i in seq_along(result$Ed_levels)) {
  level <- names(result$Ed_levels)[i]
  value <- result$Ed_levels[i]
  cat(sprintf("  %-10s Ed = %.4f\n", level, value))
}
```

**How to interpret:**

- **Species level**: Equals Shannon entropy when all species are equally
  weighted (special case where each focal element has size 1)
- **Genus level**: High when species are spread across many genera. Low when
  most species share one genus.
- **Family level**: High when genera span many families. Low when the
  community is taxonomically narrow at the family level.
- **Order level**: Similar pattern at the highest taxonomic rank.

A level with Deng entropy = 0 means **all species belong to a single group**
at that level --- it contributes no taxonomic information.

## The 8 Indices Explained

The Ozkan method produces 8 values organized in a 2 x 2 x 2 structure:

### Weighted vs Unweighted

- **Unweighted (u)**: Each taxonomic level contributes equally to the product
- **Weighted**: Higher taxonomic levels receive more weight (because resolving
  diversity at the order level is "more valuable" than at the genus level)

### With vs Without Species-Level Shannon

- **pTO**: Product of Deng entropies across taxonomic levels only
  (genus, family, order) --- pure taxonomic structure
- **pTO+**: Same product, but also includes the species-level Shannon entropy
  --- captures both abundance evenness and taxonomic structure

### All Levels vs Max-Informative Levels

- **Standard**: Uses all taxonomic levels
- **Max variants**: Uses only levels where Deng entropy > 0 (drops
  uninformative levels)

```{r eight_indices}
cat("=== All 8 Ozkan pTO indices ===\n\n")
cat("Standard (all levels):\n")
cat("  uTO      =", round(result$uTO, 4), "  (unweighted diversity)\n")
cat("  TO       =", round(result$TO, 4), "  (weighted diversity)\n")
cat("  uTO+     =", round(result$uTO_plus, 4), "  (unweighted distance)\n")
cat("  TO+      =", round(result$TO_plus, 4), "  (weighted distance)\n\n")

cat("Max-informative levels:\n")
cat("  uTO_max  =", round(result$uTO_max, 4), "  (unweighted, informative only)\n")
cat("  TO_max   =", round(result$TO_max, 4), "  (weighted, informative only)\n")
cat("  uTO+_max =", round(result$uTO_plus_max, 4), "  (unweighted distance, informative only)\n")
cat("  TO+_max  =", round(result$TO_plus_max, 4), "  (weighted distance, informative only)\n")
```

### Which index to use?

| Question | Index |
|----------|-------|
| Pure taxonomic structure (no abundance) | **uTO** or **TO** |
| Taxonomic diversity + abundance evenness | **uTO+** or **TO+** |
| Are some taxonomic levels uninformative? | Use **_max** variants |
| Default recommendation for most studies | **TO+** (most complete) |

## The Three-Run Pipeline

### Run 1: Deterministic Calculation

Uses the full community as-is. Computes all 8 indices directly.

```{r run1}
cat("Run 1 results:\n")
cat("  uTO+ =", round(result$uTO_plus, 4), "\n")
cat("  TO+  =", round(result$TO_plus, 4), "\n")
```

### Run 2: Stochastic Resampling (Slicing)

Species are removed one at a time, starting with the least abundant.
After each removal, all indices are recalculated. This "slicing" procedure
reveals two things:

1. **The maximum diversity achievable** from the community's species pool
2. **Each species' contribution** to overall diversity

```{r run2}
run2 <- ozkan_pto_resample(community, tax_tree, n_iter = 101, seed = 42)

cat("Run 1 (deterministic):  uTO+ =", round(run2$uTO_plus_det, 4), "\n")
cat("Run 2 (stochastic max): uTO+ =", round(run2$uTO_plus_max, 4), "\n")
```

**Why does maximum > deterministic?** Because some species may be
taxonomically redundant. If two species from the same genus are present,
removing one can increase the ratio of between-group to within-group
diversity. The species whose removal *increases* diversity is called
an "unhappy" species --- it is taxonomically redundant in the community.

### Visualizing Run 2

```{r run2_plot, fig.width=9, fig.height=5, fig.alt="Iteration plot showing TO+ values across stochastic resampling iterations"}
plot_iteration(run2, component = "TO_plus",
               title = "Run 2: TO+ Across Iterations")
```

**How to read:**

- **Grey dots**: pTO value for each random species subset
- **Red line**: Deterministic value (Run 1 --- all species included)
- **Blue line**: Maximum value found (Run 2 result)

Points above the red line represent subcommunities more diverse than the
full community --- evidence that some species are taxonomically redundant.

### Run 3: Max-Informative Level Variants

Some taxonomic levels carry no information. If all species belong to the
same order, Deng entropy at the order level is zero --- including it in the
product just drags the value down without adding insight.

Run 3 repeats the calculation using only levels where Deng entropy > 0:

```{r run3}
run3 <- ozkan_pto_sensitivity(community, tax_tree, run2, seed = 123)

cat("All levels:       TO+ =", round(run3$TO_plus_max, 4), "\n")
cat("Informative only: TO+ =", round(result$TO_plus_max, 4), "\n")
```

### Full Pipeline in One Call

```{r full}
full <- ozkan_pto_full(community, tax_tree, n_iter = 101, seed = 42)

cat("Complete pipeline summary:\n\n")
cat("         uTO+      TO+       uTO       TO\n")
cat("Run 1:", sprintf("%9.4f %9.4f %9.4f %9.4f",
    full$run1$uTO_plus, full$run1$TO_plus,
    full$run1$uTO, full$run1$TO), "\n")
cat("Run 2:", sprintf("%9.4f %9.4f %9.4f %9.4f",
    full$run2$uTO_plus_max, full$run2$TO_plus_max,
    full$run2$uTO_max, full$run2$TO_max), "\n")
cat("Run 3:", sprintf("%9.4f %9.4f %9.4f %9.4f",
    full$run3$uTO_plus_max, full$run3$TO_plus_max,
    full$run3$uTO_max, full$run3$TO_max), "\n")
```

## Jackknife Leave-One-Out Analysis

The jackknife procedure removes each species one at a time and recalculates
all indices. This directly measures each species' contribution:

```{r jackknife}
jk <- ozkan_pto_jackknife(community, tax_tree)

cat("Jackknife results (TO+ when each species is removed):\n\n")
jk_df <- jk$jackknife_results
for (i in seq_len(nrow(jk_df))) {
  direction <- ifelse(jk_df$TO_plus[i] > result$TO_plus, "UNHAPPY", "happy")
  cat(sprintf("  Remove %-25s -> TO+ = %.4f  [%s]\n",
              jk_df$species[i], jk_df$TO_plus[i], direction))
}

cat("\nHappy species:", jk$n_happy, "\n")
cat("Unhappy species:", jk$n_unhappy, "\n")
```

- **happy species**: Removing them *decreases* diversity (they contribute
  positively to taxonomic structure)
- **UNHAPPY species**: Removing them *increases* diversity (they are
  taxonomically redundant)

## Comparing Communities

```{r compare, fig.width=8, fig.height=8, fig.alt="Radar chart comparing diversity indices between diverse and degraded communities"}
degraded <- c(
  Quercus_coccifera = 40,
  Pinus_brutia      = 35,
  Juniperus_oxycedrus = 10
)

communities <- list(
  "Intact (10 spp)"  = community,
  "Degraded (3 spp)" = degraded
)

plot_radar(communities, tax_tree,
           title = "Intact vs Degraded Forest")
```

The radar chart reveals which diversity dimensions are most affected by
degradation. If abundance-weighted indices (Shannon, Simpson, TO+) drop
more than presence/absence indices (AvTD, uTO+), the community has lost
evenness. If both drop equally, the community has lost taxonomic breadth.

## References

- Ozkan, K. (2018). A new proposed measure for estimating taxonomic
  diversity. *Turkish Journal of Forestry*, 19(4), 336-346.
- Ozkan, K. & Mert, A. (2022). Comparisons of Deng entropy-based taxonomic
  diversity measures with the other diversity measures and introduction to
  the new proposed (reinforced) estimators. *FORESTIST*, 72(2).
  doi: [10.5152/forestist.2021.21025](https://doi.org/10.5152/forestist.2021.21025)
- Deng, Y. (2016). Deng entropy. *Chaos, Solitons & Fractals*, 91, 549-553.
- Dempster, A.P. (1967). Upper and lower probabilities induced by a
  multivalued mapping. *The Annals of Mathematical Statistics*, 38(2),
  325-339.
- Shafer, G. (1976). *A Mathematical Theory of Evidence*. Princeton
  University Press, Princeton, NJ.
